Robust receding horizon control for linear systems with state dependent disturbance

### 2015-04-21 draft_v2.tex
edited sections 4 and 5

Numerous strategies have been proposed to solve RMPC problems for linear systems subject to linear state and
input constraints, with model uncertainty in the form of bounded additive disturbances and/or multiplicative
parameters. For the case of unknown additive disturbances, \citet{Scokaert:1998} propose a scenario-based
approach to solve a sequence of min-max problems online, however the computational complexity of this approach
grows exponentially with the length of the horizon. A parametrised algorithm was proposed
by~\cite{Rakovic:2012} which reduces the RMPC problem to a single linear program to be solved online, but
although this is a closed loop strategy the parametrisation is in general suboptimal. Explicit RMPC algorithms
optimise the worst case predicted performance for each admissible model state offline and
%exploit the structure of the solution using results
use results from multi-parametric programming to parametrise the optimal control law as a piecewise affine function of the
state \citep[e.g.][]{Bemporad:2003,Diehl:2004}. However, this requires extensive offline computation and
becomes inefficient for a high-dimensional state space.
An alternative approach that similarly exploits the piecewise affine structure of the solution, but with
less redundancy, was proposed by~\cite{Buerger:2011} and extended in~\cite{Buerger:ACC}. In this work an active set
solver is designed to solve a sequence of min-max problems explicitly online.

@@

\section{Solving equality constrained multi-parametric quadratic programs}\label{sec:gen:mpQP}
In this section we discuss equality constraint multi-parametric quadratic programs (mpQP) 
(see~\cite{bemporad02,Tondel:2003} for an extended discussion), as they are essential in the treatment 
of the considered min-max program using an active set solver. Note that the notation in this section does 
not refer to the notation in the preceding and proceeding sections.

Consider the primal equality-constrained concave mpQP
\begin{equation}\label{mpQP:primal}
	V(\theta) = \left\{\begin{split}
	\max_x & \;\frac{1}{2} x^T W x + c^Tx + d\\
	\text{s.t.} &\; Hx  = f +S\theta
	\end{split}\right.
\end{equation}
for $W\preceq 0.$ The Lagrangian function associated with~\eqref{mpQP:primal} is
\begin{equation}
	L(x,\lambda,\theta)  = \frac{1}{2} x^T W x + c^Tx + d + \lambda^T(Hx - f - S\theta)
\end{equation}
and the dual problem is the unconstrained mpQP
\begin{equation}\label{mpQP:dual}\begin{split}
	\min_\lambda &\; \mathcal D(\lambda,\theta)= \min_\lambda-\frac{1}{2}\lambda^T H W^{-1}H^T \lambda 
	-(HW^{-1}c\\ &+ f + S\theta)^T\lambda
	-\frac{1}{2}c^TW^{-1}c +d
\end{split}\end{equation}
To solve~\eqref{mpQP:primal} we consider the first order optimality conditions, see e.g.~\cite{Fletcher:2000}:
\begin{equation}\label{kkt:conditions}
	\begin{aligned}
        W x + H^T\lambda & = -c \\
        H x &= f + S\theta.
	\end{aligned}
\end{equation}

\begin{thm}\label{thm:mpQP:solution}

@@

		x &= M_\theta \theta + m\\

@@

where $N_\beta\neq 0$ only if $H$ is rank deficient, in which case the rows
of $N_\beta$ form a basis of the kernel of $H$, i.e. $H^TN_\beta=0$
\end{thm}

@@

The matrices~$M_\theta$, $N_\theta$, $N_\beta$ as well as the vectors $m$ and $n$
are constant, the variable $\beta$ is used to parametrise the solution in the null space of~\eqref{kkt:conditions}.

@@

There are two cases of rank deficient $H$: rank$(H)=n_x<n_\lambda$ and
rank$(H)<\min\{n_x,n_\lambda\}$.
Using the QR-decomposition (see e.g.~\cite{Nocedal:2006}) of $H=(Q_1,Q_2)\cdot(R^T,0)^T$ the solution 

@@

@@

in the case where the number of equality constraints is larger than the dimension of
the primal variable, i.e. rank$(H)=n_x$. If rank$(H)<\min\{n_x,n_\lambda\}$, so that $R$ is
not invertible, then

@@

@@

\begin{multline}
    \mathcal D(\lambda(\theta,\beta),\theta) =  -\frac{1}{2}(N_\theta \theta + N_\beta \beta + n)^T H W^{-1}H^T 
	(N_\theta \theta\\ + N_\beta \beta + n)
	-(AW^{-1}c + 
	f + S\theta)^T(N_\theta \theta + N_\beta \beta + n)\\ -\frac{1}{2}c^TW^{-1}c,
\end{multline}

@@

and the choice of $\beta$ does not affect either the primal nor the dual cost and can be used for
any purpose. For a general mpQP~\eqref{mpQP:compatability:assumption} is a restriction on the set

@@

	V(\theta) = &\frac{1}{2}\theta^T M_\theta^T W M_\theta \theta + c^TM_\theta \theta
\\&+ m^TWM_\theta+\frac{1}{2}m^T Wm + c^Tm\\

@@

Assume now that we want to solve

@@

where the equality constraints include~\eqref{mpQP:compatability:assumption}. Since we know
that the optimal cost-to-go $V(\theta)$ has the structure~\eqref{mpQP:cost:to:go}
the problem~\eqref{mpQP:second:stage} reduces to a problem like~\eqref{mpQP:primal},
which yields a solution like~\eqref{mpQP:solution}, i.e.

@@

with the dual variable~$\eta$, as long as $\Phi+W_\theta\succ 0$. Finally we note 
that~\eqref{mpQP:compatability:assumption} is only nontrivial if~$H$ is rank deficient and only 

@@

In the solution of recursive min-max programs such as~\eqref{eq:the:problem} we use the composition of
the stage optimisers, i.e. we use $x(\theta(\phi)) = M_\theta K_\phi \phi + (M_\theta k + m)$.
Notice that this composition of optimisers is again an affine function of the parameter, in particular
we will exploit the fact that changing the parameter along a line will translate to a line in the optimiser
space as long as no activations or deactivations in the constraints occur. We will revisit this fact
when we perform a line search in order to update the active constraints.

@@

To use an active set solver for the problem:

@@

With Lemma~\ref{thm:mpQP:solution} we obtain

@@

\section{Exact solution of min-max program}\label{sec:line:search}
In the previous section we presented a recursive algorithm to solve~\eqref{eq:the:problem}
when all active constraints are known. In this section we present a method to find the active
constraints for a given initial state $\mathpzc x_e$ using a line search starting from a state $\mathpzc x_0$
for which the active constraints are known. In section~\ref{sec:gen:mpQP} and~\ref{sec:applied:mpQP}
we derived an explicit solution for each sub-optimisation program which depended affinely on the
state~$x$ at that stage~$m$ or the state and the input at that stage. Henceforth we denote the stage 
with an index, e.g.~$x_m$ denotes the optimised state in stage $m$, that is $x^{+,\ast}$ at stage $m+1$.
To ensure continuity of the dual variables we choose $\hat\zeta_m = \hat\beta_m$ and $\zeta_m=\beta_{m-1}$.
The affine character of the solution of each sub-problem allows us to parametrise the entire trajectory
with respect to parameters at stage~$m=N$, that is $x_N=\mathpzc x_0$ and $\beta_N$.

@@

\section{Example}\label{sec:example}
This section we present a numerical simulation in which the state-dependent uncertainty enters through an upper bound on the linearisation
error for a nonlinear system. We consider an inverted pendulum described by $ml^2\ddot\varphi = mlg\sin\varphi+M$, where
$m=0.1$\,kg is the mass, $l=0.3$\,m is the length of the pendulum, $g=9.81$\,m\,s$\mbox{}^{-2}$ is the gravitational
constant and $M$ is the direct torque input. Discretisation using Euler backward finite differences 
at $f = 20$\,Hz yields
\begin{equation}
  \begin{split}
    x_1[k+1] &= x_1[k] + \frac{1}{f} x_2[k]\\
    x_2[k+1] &= \frac{g}{fl} \sin(x_1[k]) + x_2[k] + \frac{1}{fml^2}u[k]
  \end{split}
\end{equation}
where $x_1[k] = \varphi(t_0+k/f),x_2[k] = \dot\varphi(t_0+k/f)$ and $u[k] = M(t_0+k/f)$.
The nonlinearity is given by the sine function, and we can explicitly state the 
linearisation error as $e(x_1)=\frac{g}{fl}(\sin x_1 - x_1)$. Since the linearisation
only approximates the nonlinear dynamics around the equilibrium we can choose a closed interval 
$0\in D\subset\mathbb R$ and maximise the upper bound 
\begin{equation}
  \Delta e = \max_{x_1\in D}\abs{\frac{de}{dx_1}} = \frac{g}{fl}\max_{x_1\in D}\abs{\cos x_1-1}.
\end{equation}
This results in the linear system
\begin{equation}
  x^+ = \underbrace{\left(\begin{array}{cc}
  1 & f^{-1}\\ g(fl)^{-1} & 1
  \end{array}\right)}_A x + \underbrace{\left(\begin{array}{c} 0 \\(ml^2f)^{-1} \end{array}\right)}_B u
  +\left(\begin{array}{c}w_1\\ w_2\end{array}\right)
\end{equation}
with $\abs{w_1}\leq w_{1,\max}$ and $\abs{w_2}\leq\max\{w_{2,\max},\Delta e \abs{x_1}\}$.
We also introduce the input constraint $\abs{u}\leq3$\,N\,m and use the interval $D = [-\frac{3\pi}{4},
\frac{3\pi}{4}]$ and a horizon of $N=7$ to obtain the sequence of state constraints $\mathcal X_m$
depicted in Figure~\ref{fig:numerical:example}. The line search starting at the origin and
exploring towards $\mathpzc x_e=(-6,50)$ produces the optimal trajectories at the active-set switches
which are illustrated in blue. In pink a warm-started line search terminates on the boundary of the
feasible set before reaching the (infeasible) state $\mathpzc x_e=(0,30)$.

>>>>

Various strategies have been proposed to solve RMPC problems for linear systems subject to linear state and
input constraints, with model uncertainty in the form of bounded additive disturbances and/or multiplicative
parameters. For the case of unknown additive disturbances, \citet{Scokaert:1998} propose a scenario-based
approach to solve a sequence of min-max problems online, however the computational complexity of this approach
grows exponentially with the length of the horizon. A parametrised algorithm was proposed
by~\cite{Rakovic:2012} which reduces the RMPC problem to a single linear program to be solved online, but
although this is a closed loop strategy the parametrisation is in general suboptimal. Explicit RMPC algorithms
optimise the worst case predicted performance for each admissible model state offline and
%exploit the structure of the solution using results
use results from multi-parametric programming to parametrise the optimal control law as a piecewise affine function of the
state \citep[e.g.][]{Bemporad:2003,Diehl:2004}. However, this requires extensive offline computation and
becomes inefficient for a high-dimensional state space.
An alternative approach that similarly exploits the piecewise affine structure of the solution, but with
less redundancy, was proposed by~\cite{Buerger:2011} and extended in~\cite{Buerger:ACC}. In this work an active set
solver is designed to solve a sequence of min-max problems explicitly online.

@@

\section{Solving equality constrained multi-parametric quadratic programs}\label{sec:gen:mpQP}
In this section we discuss equality constraint multi-parametric quadratic programs (mpQP) 
(see~\cite{bemporad02,Tondel:2003} for an extended discussion), as they are essential in the treatment 
of the considered min-max program using an active set solver. 
%Note that the notation in this section does not refer to the notation in the preceding and proceeding sections.
Consider the primal equality-constrained concave mpQP
\begin{equation}\label{mpQP:primal}
	V(\theta) = \left\{\begin{split}
	\max_{y} & \;\frac{1}{2} y^T W y + c^Ty + d\\
	\text{s.t.} &\; Hy  = f +S\theta
	\end{split}\right.
\end{equation}
%where  $y\in\mathbb{R}^{n_y}$ is the optimization variable, 
where $W\preceq 0$, $c$, $d$, $H$, $f$, $S$ are constants and $\theta$ is a variable parameter. We wish to
know how the solution of (\ref{mpQP:primal}) varies with $\theta$.  The Lagrangian function
associated with~\eqref{mpQP:primal} is
\[
	L(y,\lambda,\theta)  = \frac{1}{2} y^T W y + c^Ty + d + \lambda^T(Hy - f - S\theta)
\]
and the dual problem is the unconstrained mpQP
\begin{equation}\label{mpQP:dual}\begin{split}
	\min_\lambda &\; \mathcal D(\lambda,\theta)= \min_\lambda-\frac{1}{2}\lambda^T H W^{-1}H^T \lambda 
	-(HW^{-1}c\\ &+ f + S\theta)^T\lambda
	-\frac{1}{2}c^TW^{-1}c +d
\end{split}\end{equation}
To solve~\eqref{mpQP:primal} we consider the first order optimality conditions, see e.g.~\cite{Fletcher:2000}:
\begin{equation}\label{kkt:conditions}
	\begin{aligned}
        W y + H^T\lambda & = -c \\
        H y &= f + S\theta.
	\end{aligned}
\end{equation}

\begin{lem}\label{lem:mpQP:solution}

@@

		y &= M_\theta \theta + m\\

@@

where $N_\beta\neq 0$ only if $H$ is not right invertible, in which case $H^TN_\beta=0$ and the columns
of $N_\beta$ span ker$(H^T)$.
\end{lem}

@@

The matrices $M_\theta$, $N_\theta$, $N_\beta$ and vectors $m$, $n$
are constant and the variable $\beta$ is used to parametrise the solution in the null space
of~\eqref{kkt:conditions}. We say (\ref{mpQP:primal}) is \textit{degenerate} if $N_\beta\neq 0$.

@@

If $H$ is not right invertible, then we must have rank$(H)=n_y<n_\lambda$ or
rank$(H)<\min\{n_y,n_\lambda\}$.
For the case that rank$(H)=n_y$ (i.e.\ $H$ has full rank and $n_y<n_\lambda$), 
using the QR-decomposition \citep[e.g.][]{Nocedal:2006} of $H=(Q_1,Q_2)\cdot(R^T,0)^T$ the solution 

@@

\[

@@

	\end{array}\right) .
\]
On the other hand if rank$(H)<\min\{n_y,n_\lambda\}$ (so that $H$ is rank deficient), then
\[

@@

\]

@@

\begin{align*}
\mathcal D(&\lambda(\theta,\beta),\theta) = \\
& {-\tfrac{1}{2}}(N_\theta \theta + N_\beta \beta + n)^T H W^{-1}H^T 
	(N_\theta \theta + N_\beta \beta + n)\\
& -(AW^{-1}c + 
	f + S\theta)^T(N_\theta \theta + N_\beta \beta + n) -\tfrac{1}{2}c^TW^{-1}c,
\end{align*}

@@

and the choice of $\beta$ does not affect either the primal or the dual cost and can be used for
any purpose. For a general mpQP, condition~\eqref{mpQP:compatability:assumption} is a restriction on the set

@@

	V(\theta) = &\frac{1}{2}\theta^T M_\theta^T W M_\theta \theta + c^TM_\theta \theta \\
& + m^TWM_\theta+\frac{1}{2}m^T Wm + c^Tm\\

@@

Suppose now that we want to solve a problem that depends on the solution of~\eqref{mpQP:primal} and an
additional parameter $\phi$:

@@

where $\Phi$, $C$, $g$, $f$, $C$, $e$, $T$ are constants and
the equality constraints include~\eqref{mpQP:compatability:assumption}. Since we know
that the optimal cost-to-go $V(\theta)$ has the structure~\eqref{mpQP:cost:to:go},
the problem~\eqref{mpQP:second:stage} reduces to a problem with the form of~\eqref{mpQP:primal},
which yields a solution similar to~\eqref{mpQP:solution}, i.e.

@@

where $\eta$ is the dual variable, and $\Phi+W_\theta\succ 0$ has been assumed. Finally we note 
that~\eqref{mpQP:compatability:assumption} is only nontrivial if~$H$ is not right invertible and only 

@@

In the solution of recursive min-max programs such as~\eqref{eq:the:problem} we use the composition of
the stage optimisers, i.e. we use $y(\theta(\phi)) = M_\theta K_\phi \phi + (M_\theta k + m)$.
Notice that this composition of optimisers is again an affine function of the parameter $\phi$, in particular
we will exploit the fact that parameter variations along lines translate to lines in the optimiser
space as long as no activations or deactivations of constraints occur. We will revisit this fact
when we perform a line search in order to update the active constraints.

@@

%To use an active set solver for the problem:
First epx

@@

With Lemma~\ref{lem:mpQP:solution} we obtain

@@

\section{Exact solution of min-max program}\label{sec:line:search}
In the previous section we presented a recursive algorithm to solve~\eqref{eq:the:problem}
when all active constraints are known. In this section we present a method to find the active
constraints for a given initial state $\mathpzc x_e$ using a line search starting from a state $\mathpzc x_0$
for which the active constraints are known. In section~\ref{sec:gen:mpQP} and~\ref{sec:applied:mpQP}
we derived an explicit solution for each sub-optimisation program which depended affinely on the
state~$x$ at that stage~$m$ or the state and the input at that stage. Henceforth we denote the stage 
with an index, e.g.~$x_m$ denotes the optimised state in stage $m$, that is $x^{+,\ast}$ at stage $m+1$.
To ensure continuity of the dual variables we choose $\hat\zeta_m = \hat\beta_m$ and $\zeta_m=\beta_{m-1}$.
The affine character of the solution of each sub-problem allows us to parametrise the entire trajectory
with respect to parameters at stage~$m=N$, that is $x_N=\chi_0$ $x_N=\mathpzc x_0$ and $\beta_N$.

@@

\input{example.tex}
